Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent

### 2023-06-28 manuscript/main.tex
more paragraph 2

	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in a HOGWILD! parallel manner.
	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
	Our implementation produces layouts in 1 dimension (1D) and 2 dimensions (2D), but it can be extended in any number of dimensions.

@@

	%In this work, we describe the results in 1 dimension (1D) and 2 dimensions (2D). In this section we will describe. Then we will. And then. \\

\begin{definition}
	Variation graphs are a mathematical formalism to represent pangenome graphs \citep{Garrison_2019_thesis}.
	In the variation graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{P})$, nodes (or vertices) $\mathcal{V} = v_1\ldots v_{|\mathcal{V}|}$ contain sequences of DNA.
	Each node $v_i$ has a unique identifier $i$ and an implicit reverse complement $\bar{v_i}$.
	The node strand $s$ represents node orientation.
	Edges $\mathcal{E} = e_1\ldots e_{|\mathcal{E}|}$ connect ordered pairs of node strands ($e_i = ( s_a, s_b )$), defining the graph topology.
	Paths $\mathcal{P} = p_1\ldots p_{|\mathcal{P}|}$ describe walks over node strands ($p_i = s_1 \ldots s_{|p_i|}$), representing the genomes embedded in the graph.
\end{definition}

	Our HOGHWILD! path-guided stochastic gradient descent algorithm is inspired by \cite{Zheng2019} and is theoretically applicable in any number of dimensions. 
	We implemented the algorithm in ODGI \citep{Guarracino2022}, making use of its already proven interface to pangenome graphs in the standard graphical fragment assebmly (GFA) format (\url{https://github.com/GFA-spec/GFA-spec}).
	In our implementation, the algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the actual nucleotide distance of a path traversing these nodes.
	Our path index enables efficient navigation of path positions by updating nodes on the fly, instead of keeping all pairwise distances of all nodes in memory.
	This allows the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
	Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
	\begin{itemize}
	    \item The first node Xi of a pair is a uniform path step pick from all nodes.
	    \item The second node Xj of a pair is sampled from the same path following a Zipfian distribution.
	    \item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes.
	    \item The magnitude r of the update depends on the current learning rate of the SGD.
	\end{itemize}
	%The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.
	We describe in algorithm~\ref{alg:pg_sgd} how the PG-SGD works on a single thread.
	We can have several worker threads but only one checker thread!

	\subsection{general algorithm}
	\subsection{path index}
	\subsection{parallelization in detail}
	\subsection{randomization}
	ZIPF!
	\subsection{annealing step-size $\eta$}
	\subsection{convergence / stop criteria, iterations, term updates}
	\subsection{cooling || flip}
	\subsection{from layout to 1D node order}

>>>>

	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to move pairs of nodes in a HOGWILD! parallel manner.
	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
	It can be extended in any number of dimensions. We provide in the ODGI toolkit \citep{Guarracino2022} implementations to produce layouts in 1 dimension (1D) and 2 dimensions (2D).

@@

	The PG-SGD algorithm is inspired by \cite{Zheng2019}, but we designed it to work on pangenomes in the form of variation graph (\ref{def:vg}).

\begin{definition}
	\label{def:vg}
	Variation graphs are a mathematical formalism to represent pangenome graphs \citep{Garrison_2019_thesis}.
	In the variation graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{P})$, nodes (or vertices) $\mathcal{V} = v_1\ldots v_{|\mathcal{V}|}$ contain nucleotide sequences.
	Each node $v_i$ has a unique identifier $i$ and an implicit reverse complement $\bar{v_i}$.
	The node strand $s$ represents node orientation.
	Edges $\mathcal{E} = e_1\ldots e_{|\mathcal{E}|}$ connect ordered pairs of node strands ($e_i = ( s_a, s_b )$), defining the graph topology.
	Paths $\mathcal{P} = p_1\ldots p_{|\mathcal{P}|}$ describe walks over node strands ($p_i = s_1 \ldots s_{|p_i|}$), representing the genomes embedded in the graph.
\end{definition}

	The algorithm moves a single pair of nodes at a time, minimizing the disparity between the layout distance of a node pair and the nucleotide distance of the same nodes as calculated along a path that traverses them.
	Our path index enables efficient retrieval of path positions to quickly compute nucleotide distances.
	This eliminates the need to store all pairwise distances between nodes in memory, allowing the PG-SGD to be applied to gigabase-scale pangenome graphs. \\
	%Our multithreaded implementation (\url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}) presents a working prototype that is based on succinct graph data structures (\url{https://doi.org/ghqdzw}):
	Our implementation is multithreaded, but for the sake of simplicity, we describe how the algorithm works on a single thread (~\ref{alg:pg_sgd}).

	\subsection{General algorithm}

	\FIXME{
		\begin{itemize}
			\item The first node Xi of a pair is a uniform path step pick from all nodes.
			\item The second node Xj of a pair is sampled from the same path following a Zipfian distribution.
			\item The path nucleotide distance of the nodes in the pair guides the actual layout distance dij update of these nodes.
			\item The magnitude r of the update depends on the current learning rate of the SGD.
		\end{itemize}
	}
	%The exploration of the path-guided SGD parameter space is in progress, in order to get the best layout as quickly as possible.

	\subsection{Path index}
	%\subsection{parallelization in detail}
	\subsection{Layoit initialization}
	\subsection{Zipfian distribution}
	\subsection{Annealing}
	\subsection{Stop criteria, iterations, term updates}
	\subsection{Cooling and flip}
	\subsection{From 1D layout to 1D node order}
